DATA DRIVEN MODELS FOR APOGEE M DWARFS

### 2018-04-12 paper/draft1/draft1.tex
add cannon eqns

\usepackage{color}
\usepackage{amsmath}
\usepackage{enumitem}
\usepackage{graphicx}

\definecolor{bcolor}{RGB}{0, 51, 153}
\hypersetup{linkcolor=bcolor,citecolor=bcolor,filecolor=cyan,urlcolor=bcolor}

@@

\author[0000-0002-7961-6881]{Jessica Birky}
\affil{Center for Astrophysics and Space Science, University of California San Diego, La Jolla, CA 92093, USA}
\affil{Max-Planck-Institut f\"ur Astronomie, K\"onigstuhl 17, D-69117 Heidelberg}

\author[0000-0003-2866-9403]{David W. Hogg}
\affil{Max-Planck-Institut f\"ur Astronomie, K\"onigstuhl 17, D-69117 Heidelberg}
\affiliation{Center for Computational Astrophysics, Flatiron Institute, 162 Fifth Ave, New York, NY 10010, USA}
\affiliation{Center for Cosmology and Particle Physics, Department of Physics, New York University, 726
Broadway, New York, NY 10003, USA}
\affil{Center for Data Science, New York University, 60 Fifth Ave, New York, NY 10011, USA}

@@

@@

The Cannon \citep{Ness:2015} is a flexible, data-driven spectral modeling and parameter inference framework, demonstrated on high-resolution Apache Point Galactic Evolution Experiment (APOGEE; $\lambda/\Delta\lambda\sim22,500$, $1.5-1.7 \mu m$) spectra of giant stars to estimate stellar labels (T$_{\rm eff}$, logg, [Fe/H], and chemical abundances) to precisions higher than the model-grid pipeline. The lack of reliable stellar parameters reported by the APOGEE pipeline for temperatures less than $\sim3550K$ \citep{Schmidt:2016}, motivates the extension of this approach to M dwarf stars. Using a training set of 51 M dwarfs with spectral types ranging M0-M9 obtained from SDSS optical spectra, we demonstrate that The Cannon can infer spectral types to a precision of 0.6 types. We then use 30 M dwarfs ranging $3072 < T_{\rm eff} < 4131K$, and -0.48 $<$ [Fe/H] $<$ 0.49 to train a two-parameter model precise to 44K and 0.05 dex respectively. Additionally we discuss the extension of a model to other labels, and the scientific objectives a data-driven pipeline could enable.

@@

% \begin{enumerate}
% \item[-] Why M-dwarfs are important: most abundant type of star, are planet hosts (The CARMENES search for exoplanets around M dwarfs \citealt{Trifonov:2018}; some famous examples: Trappist 1 (M8) -- \citealt{Gillon:2017}; Kepler-186, Kepler-138... other examples?), have lifespans longer than the age of the universe (billions or up to trillions of years) \citep{Laughlin:1997} making them an interesting probe of Milky way populations \citep{Bochanski:2007}.
% \end{enumerate}
Very low mass (VLM) are the most abundant type of star, comprising $\sim 70 \%$ of the galaxy's population by number \citep{Bochanski:2010} and $\sim 40 \%$ by mass are known to have lifespans longer than the age of the universe (billions or up to trillions of years) \citep{Laughlin:1997} making them an interesting probe of Milky way populations \citep{Bochanski:2007}.

@@

The physical parameter sample consists of 30 sources with reference labels distributed over $3072 < T_{\rm eff} < 4131K$, and -0.48 $<$ [Fe/H] $<$ 0.49 (Figure \ref{fig:train_dist}) drawn from \citealt{Mann:2015} (hereafter M15). The M15 catalog in total contains 183 sources selected from CONCH-SHELL \citep{Gaidos:2013} and with infrared spectra taken in follow up from the SpeX Spectograph. Effective temperatures are measured by comparison of BT-SETTL atmospheric models \citep{Allard:2011} to optical spectra. A subsample of 29 sources with measured angular diameters and bolometric fluxes from long-baseline optical interferometry are used to callibrate effective temperature measurements. Relative uncertainty from the callibration sample is measured to be \color{red}{$\sim$30K (check this)}\color{black}, which added in quadrature to the scatter from model comparison gives a total reported uncertainty of $\sim$60K in T$_{\rm eff}$.

@@

\color{red}{why logg not included as atmospheric parameter?}\color{black}

@@

APOGEE data products come in several types of files: ap1D, apVisit, apStar, and aspcapStar. SNR of stars in our sample.

@@

\begin{enumerate}
\item[-] A few-paragraph summary of what The Cannon does, with references (\citealt{Ness:2015}, \citealt{Casey:2016}, \citealt{Ho:2017a}, \citealt{Ho:2017b})
\end{enumerate}

@@

\subsection{Data-Driven Model}
\begin{enumerate}
\item[-] Describe assumptions of the model: (1) Sources with identical labels have near-identical flux at each pixel; (2) Expected flux at each pixel varies continuosly with change in label. 
\item[-] Briefly describe modeling procedure (training step, test step, quadratic labels, etc.), refer to \citealt{Ness:2015} for full description.
\item[-] Briefly describe model inputs and outputs (1D spectral type model and 2D teff/metallicity and label vectors for those). Mention other models we've tested.
\end{enumerate}

@@

\begin{figure}[ht]
\plotone{figures/1_physical_model.png}
\plotone{figures/2_physical_model.png}
\plotone{figures/3_physical_model.png}
\caption{\textit{Top two panels:} Mann-trained model for varying temperatures; \textit{bottom two panels:} Mann-trained model for varying metallicities.} \label{fig:mann_demo}
\end{figure}

@@

\item[-] Some discussion of precision and accuracy: The Cannon can produce precise results, but is only as accurate as its training set.

@@

\item[-] Comment on specific features? \color{red}{Can easily plot some line lists, but not sure where to go with the analysis... What features are important for M dwarfs? Compare derivative plots of specific lines between theory-trained models and data-trained models?}\color{black}

@@

%%==========================================================================================
\acknowledgements
We would like to acknowledge Derek Homeier (Universit{\"a}t Heidelberg), Wolfgang Bradner (MPIA), Hans Walter Rix (MPIA), Anna-Christina Eilers (MPIA), Anna Ho (Caltech), and Jackie Faherty (AMNH) for constructive discussions in the process of this project, as well as Bertrand Goldman (MPIA) and those who have supported the internship program at the Max Planck Institute f{\"u}r Astronomie for providing JB with funding and hospitality. Software: The Cannon, numpy, matplotlib...

\color{red}{HOGG: SDSS acknowledgments... Grants and stuff...}\color{black}

>>>>

\usepackage{color}
\usepackage{amsmath}
\usepackage{enumitem}
\usepackage{graphicx}
\usepackage{mathrsfs}
\usepackage[mathscr]{euscript}

\definecolor{bcolor}{RGB}{0, 51, 153}
\definecolor{gcolor}{RGB}{51, 153, 51}
\hypersetup{linkcolor=bcolor,citecolor=bcolor,filecolor=cyan,urlcolor=bcolor}

@@

\author[0000-0002-7961-6881]{Jessica Birky}
\affil{Center for Astrophysics and Space Science, University of California San Diego, La Jolla, CA 92093, USA}
\affil{Max-Planck-Institut f\"ur Astronomie, K\"onigstuhl 17, D-69117 Heidelberg, Germany}

\author[0000-0003-2866-9403]{David W. Hogg}
\affil{Max-Planck-Institut f\"ur Astronomie, K\"onigstuhl 17, D-69117 Heidelberg, Germany}
\affiliation{Center for Computational Astrophysics, Flatiron Institute, 162 Fifth Ave, New York, NY 10010, USA}
\affiliation{Center for Cosmology and Particle Physics, Department of Physics, New York University, 726
Broadway, New York, NY 10003, USA}
\affil{Center for Data Science, New York University, 60 Fifth Ave, New York, NY 10011, USA}

@@

\author[0000-0002-6523-9536]{Adam Burgasser}
\affil{Center for Astrophysics and Space Science, University of California San Diego, La Jolla, CA 92093, USA}

@@

The Cannon \citep{Ness:2015} is a flexible, data-driven spectral modeling and parameter inference framework, demonstrated on high-resolution Apache Point Galactic Evolution Experiment (APOGEE; $\lambda/\Delta\lambda\sim22,500$, $1.5-1.7 \mu m$) spectra of giant stars to estimate stellar labels (T$_{\rm eff}$, logg, [Fe/H], and chemical abundances) to precisions higher than the model-grid pipeline. The lack of reliable stellar parameters reported by the APOGEE pipeline for temperatures less than $\sim3550K$ \citep{Schmidt:2016}, motivates the extension of this approach to M dwarf stars. Using a training set of 51 M dwarfs with spectral types ranging M0-M9 obtained from SDSS optical spectra, we demonstrate that The Cannon can infer spectral types to a precision of 0.6 types. We then use 30 M dwarfs ranging $3072 < T_{\rm eff} < 4131K$, and -0.48 $<$ [Fe/H] $<$ 0.49 to train a two-parameter model precise to 44K and 0.05 dex respectively. Finally we demonstrate that the two-parameter Cannon model replicates the spectral features (\color{gcolor}{HOGG:}\color{black} report some pixel accracy?)

to theoretical models of M dwarf atmospheres (PHOENIX, \citealt{Husser:2013}; BTSETTL, \citealt{Allard:2011}) and identify features of missing line strengths.

@@

% \begin{enumerate}
% \item[-] Why M-dwarfs are important: most abundant type of star, are planet hosts (The CARMENES search for exoplanets around M dwarfs \citealt{Trifonov:2018}; some famous examples: Trappist 1 (M8) -- \citealt{Gillon:2017}; Kepler-186, Kepler-138... other examples?), have lifespans longer than the age of the universe (billions or up to trillions of years) \citep{Laughlin:1997} making them an interesting probe of Milky way populations \citep{Bochanski:2007}.
% \end{enumerate}
Very low mass (VLM) stars are the most abundant type of star, comprising $\sim 70 \%$ of the galaxy's population by number \citep{Bochanski:2010} and $\sim 40 \%$ by mass are known to have lifespans longer than the age of the universe (billions or up to trillions of years) \citep{Laughlin:1997} making them an interesting probe of Milky way populations \citep{Bochanski:2007}.

@@

The physical parameter sample consists of 30 sources with reference labels distributed over $3072 < T_{\rm eff} < 4131K$, and -0.48 $<$ [Fe/H] $<$ 0.49 (Figure \ref{fig:train_dist}) drawn from \citealt{Mann:2015} (hereafter M15). The M15 catalog in total contains 183 sources selected from CONCH-SHELL \citep{Gaidos:2013} and with infrared spectra taken in follow up from the SpeX Spectograph. Effective temperatures are measured by comparison of BT-SETTL atmospheric models \citep{Allard:2011} to optical spectra. A subsample of 29 sources with measured angular diameters and bolometric fluxes from long-baseline optical interferometry are used to callibrate effective temperature measurements. Relative uncertainty from the callibration sample is measured to be \color{red}{$\sim$30K (check this)}\color{black}, which added in quadrature to the scatter from model comparison gives a total reported uncertainty of $\sim$60K in T$_{\rm eff}$. Metallicities are measured

@@

We note that surface gravity (logg) and rotational velocity (vsini) parameters are not included as training labels because...

@@

APOGEE data products come in several types of files: ap1D, apVisit, apStar, and aspcapStar. SNR of stars in our sample \citealt{Perez:2016}.

@@

\begin{enumerate}
\item[-] A few-paragraph summary of what The Cannon does, with references (\citealt{Ness:2015}, \citealt{Casey:2016}, \citealt{Ho:2017a}, \citealt{Ho:2017b})
\item[-] Describe assumptions of the model: (1) Sources with identical labels have near-identical flux at each pixel; (2) Expected flux at each pixel varies continuosly with change in label. 
\item[-] Briefly describe modeling procedure (training step, test step, quadratic labels, etc.), refer to \citealt{Ness:2015} for full description.
\item[-] Briefly describe model inputs and outputs (1D spectral type model and 2D teff/metallicity and label vectors for those). Mention other models we've tested.
\end{enumerate}

@@

Following from \citealt{Ness:2015} and \citealt{Ho:2017a} 
\begin{equation}
	f_{n\lambda} = g(\ell_n | \theta_{\lambda}) + {\rm noise}
\end{equation}

\begin{equation}
	f_{n\lambda} = \theta_{\lambda}^{T} \cdot \ell_{n} + [s_{\lambda}^{2} + \sigma_{n\lambda}^{2}] \,\xi_{n\lambda}
\end{equation}

Quadratic label vectors:
\begin{equation}
	\ell_{n} = [1, SPT, SPT^{2}]
\end{equation}

\begin{equation}
	\ell_{n} = [1, T_{\rm eff}, [Fe/H], T_{\rm eff} \cdot [Fe/H], [Fe/H]^{2}]
\end{equation}

@@

\begin{figure}[ht]
\plotone{figures/1_physical_model.png}
\plotone{figures/2_physical_model.png}
\plotone{figures/3_physical_model.png}
\caption{\textit{Top two panels:} Mann-trained model for varying temperatures; \textit{bottom two panels:} Mann-trained model for varying metallicities. \color{red}{Things to add: line list and model - data difference}\color{black}} \label{fig:mann_demo}
\end{figure}

\begin{enumerate}
\item[-] Describe derivative plots and jacknife error estimation.
\end{enumerate}

@@

\item[-] \color{gcolor}{HOGG: Some discussion of precision and accuracy: The Cannon can produce precise results, but is only as accurate as its training set.}\color{black}

@@

@@

%%==========================================================================================
\acknowledgements
We would like to acknowledge Derek Homeier (Universit{\"a}t Heidelberg), Wolfgang Bradner (MPIA), Hans Walter Rix (MPIA), Anna-Christina Eilers (MPIA), Anna Ho (Caltech), and Jackie Faherty (AMNH) for constructive discussions in the process of this project, as well as Bertrand Goldman (MPIA) and those who have supported the internship program at the Max Planck Institute f{\"u}r Astronomie for providing JB with funding and hospitality. \color{red}Software: The Cannon, numpy, matplotlib... \color{black}

\color{gcolor}{HOGG: SDSS acknowledgments... Grants and stuff...}\color{black}
